---
title: "Multivariate Comparison of Two Biting-Fly Species"
subtitle: "Statistics 232C - Midterm Take-home, Spring 2025"
author: "Author 1 & Author 2"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    keep_tex: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: cosmo
fontsize: 11pt
geometry: margin=1in
header-includes:
  - \usepackage{booktabs}
  - \usepackage{float}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = FALSE,    # hide code in the body; reprinted in appendix
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  fig.width  = 6.5,
  fig.height = 4.2,
  out.width  = "85%"
)

# ---- Required packages -----------------------------------------------------
pkgs <- c("readxl", "dplyr", "tidyr", "ggplot2", "GGally",
          "MVN", "biotools", "ICSNP", "MASS", "knitr", "kableExtra")
new  <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new)) install.packages(new, repos = "https://cloud.r-project.org")
invisible(lapply(pkgs, library, character.only = TRUE))

set.seed(2025)
```

# Introduction

Two species of biting flies of the genus *Leptoconops* — *L. torrens* and
*L. carteri* — are nearly indistinguishable in external morphology and were
long believed to be a single species. Subsequent biological studies revealed
differences in sex ratio at emergence and in biting behaviour. The data set
`flies.xls` records seven morphometric measurements taken on 35 flies of each
species (total $n = 70$). The aim of this analysis is to determine, using
multivariate statistical methodology, whether the two populations differ and,
if so, on which morphological characters.

The seven measured variables are:

| Code | Variable | Meaning |
|------|----------|---------|
| $X_1$ | `W_len`   | Wing length |
| $X_2$ | `W_wid`   | Wing width  |
| $X_3$ | `P3_len`  | Third-palp length |
| $X_4$ | `P3_wid`  | Third-palp width  |
| $X_5$ | `P4_len`  | Fourth-palp length |
| $X_6$ | `Ant12`   | Length of antennal segment 12 |
| $X_7$ | `Ant13`   | Length of antennal segment 13 |

The response indicator `Species` equals 0 for *L. torrens* and 1 for
*L. carteri*.

# Data

```{r load-data}
df <- read_excel("flies.xls")
names(df) <- c("W_len", "W_wid", "P3_len", "P3_wid",
               "P4_len", "Ant12", "Ant13", "Species")
df <- df %>%
  mutate(Species = factor(Species, levels = c(0, 1),
                          labels = c("L.torrens", "L.carteri")))
glimpse(df)
```

```{r split}
X  <- as.matrix(df[, 1:7])
g  <- df$Species
X1 <- X[g == "L.torrens", ]   # n1 = 35
X2 <- X[g == "L.carteri", ]   # n2 = 35
n1 <- nrow(X1); n2 <- nrow(X2); p <- ncol(X)
```

## Descriptive statistics

```{r desc-tbl}
desc <- df %>%
  pivot_longer(-Species, names_to = "Variable", values_to = "Value") %>%
  group_by(Variable, Species) %>%
  summarise(Mean = mean(Value), SD = sd(Value), .groups = "drop") %>%
  pivot_wider(names_from = Species, values_from = c(Mean, SD)) %>%
  mutate(Diff = Mean_L.torrens - Mean_L.carteri)

kable(desc, digits = 2, booktabs = TRUE,
      caption = "Per-species means and standard deviations.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r boxplot, fig.cap="Side-by-side boxplots of the seven variables."}
df %>%
  pivot_longer(-Species, names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(Species, Value, fill = Species)) +
  geom_boxplot(alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free_y", ncol = 4) +
  theme_bw() + theme(legend.position = "none")
```

```{r pairs, fig.cap="Pair-wise scatter plot matrix coloured by species.", fig.height=6}
ggpairs(df, columns = 1:7, aes(colour = Species, alpha = 0.6),
        upper = list(continuous = wrap("cor", size = 2.6))) +
  theme_bw(base_size = 8)
```

We first carry out exploratory data analysis. From the table and plots above, we can learn that *L. carteri* tends to have larger measurements, especially in wing length, third palp length and fourth palp length. Moreover, the boxplots indicate that there might exist a few outliers and pairwise scatterplots show that the data is not in a Gaussian distribution in general.

# Normality assessment

## Univariate Q–Q plots and Shapiro–Wilk tests

```{r qq-univ, fig.cap="Normal Q–Q plots for each variable, by species.", fig.height=7}
op <- par(mfrow = c(7, 2), mar = c(3, 3, 2, 1), mgp = c(1.6, 0.5, 0))
for (k in 1:p) {
  qqnorm(X1[, k], main = paste(colnames(X)[k], "  L.torrens"))
  qqline(X1[, k], col = "red")
  qqnorm(X2[, k], main = paste(colnames(X)[k], "  L.carteri"))
  qqline(X2[, k], col = "red")
}
par(op)
```

```{r shapiro}
shap <- sapply(1:p, function(k) {
  c(L.torrens = shapiro.test(X1[, k])$p.value,
    L.carteri = shapiro.test(X2[, k])$p.value)
})
colnames(shap) <- colnames(X)
kable(t(round(shap, 3)), booktabs = TRUE,
      caption = "Shapiro–Wilk p-values (variables in rows).") %>%
  kable_styling(latex_options = "hold_position")
```

To check the normality for this data set, we first check the univariate normality and then the multivariate normality. The qqplots and Shapiro–Wilk tests for each variables indicate that almost every variable violates the normality assumption.

## Multivariate normality

```{r mvn}
# Self-contained Mardia test: avoids MVN 6.x output formatting that
# breaks LaTeX (Unicode glyphs, special chars).
mardia_test <- function(X) {
  X <- as.matrix(X); n <- nrow(X); p <- ncol(X)
  Xc <- scale(X, center = TRUE, scale = FALSE)
  S  <- cov(X) * (n - 1) / n
  D  <- Xc %*% solve(S) %*% t(Xc)

  b1p <- sum(D^3) / n^2
  b2p <- sum(diag(D)^2) / n
  k1  <- n * b1p / 6
  df1 <- p * (p + 1) * (p + 2) / 6
  p1  <- pchisq(k1, df1, lower.tail = FALSE)
  z2  <- (b2p - p * (p + 2)) / sqrt(8 * p * (p + 2) / n)
  p2  <- 2 * pnorm(-abs(z2))

  data.frame(
    Test       = c("Mardia skewness", "Mardia kurtosis"),
    Statistic  = round(c(k1, z2), 3),
    df         = c(df1, NA),
    p_value    = signif(c(p1, p2), 3),
    Reject_H0  = ifelse(c(p1, p2) < 0.05, "Yes", "No"),
    stringsAsFactors = FALSE
  )
}

kable(mardia_test(X1), booktabs = TRUE,
      caption = "Mardia test of multivariate normality, L. torrens.") %>%
  kable_styling(latex_options = "hold_position")
kable(mardia_test(X2), booktabs = TRUE,
      caption = "Mardia test of multivariate normality, L. carteri.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r chisq-qq, fig.cap="Chi-squared Q–Q plots of squared Mahalanobis distances."}
chiQQ <- function(M, ttl) {
  d2   <- mahalanobis(M, colMeans(M), cov(M))
  qchi <- qchisq(ppoints(nrow(M)), df = ncol(M))
  plot(qchi, sort(d2), pch = 20, main = ttl,
       xlab = expression(chi[7]^2 ~ "quantile"),
       ylab = "ordered  d^2")
  abline(0, 1, col = "red", lwd = 1.5)
}
op <- par(mfrow = c(1, 2))
chiQQ(X1, "L. torrens"); chiQQ(X2, "L. carteri"); par(op)
```

To check the multivariate normality, the qqplot compares ordered Mahalanobis distances with theoretical chi-square quantiles. As shown in the plots, some points lie far away from the straight line, so the multivariate normality assumption is unreasonable. Therefore, we might need to transform the raw data.

# Outlier detection

```{r outliers}
cutoff <- qchisq(0.975, df = p)        # ~ 16.01
d2_1   <- mahalanobis(X1, colMeans(X1), cov(X1))
d2_2   <- mahalanobis(X2, colMeans(X2), cov(X2))

out_tbl <- data.frame(
  Species  = c(rep("L.torrens", n1), rep("L.carteri", n2)),
  Index    = c(seq_len(n1), seq_len(n2)),
  D2       = c(d2_1, d2_2),
  Outlier  = c(d2_1, d2_2) > cutoff
)
kable(subset(out_tbl, Outlier),
      digits = 2, booktabs = TRUE,
      caption = sprintf("Cases with $d^2 > %.2f$.", cutoff)) %>%
  kable_styling(latex_options = "hold_position")
```

# Transformations

To stabilise variance and improve approximate joint normality, we
estimated the optimal Box-Cox parameter $\lambda$ separately for each of
the seven morphometric variables (using the pooled sample, so the same
$\lambda$ is applied to both species and the variables remain on
comparable scales).

```{r boxcox, results='hide', fig.show='hide'}
lambda <- sapply(1:p, function(k) {
  bc <- MASS::boxcox(X[, k] ~ 1, plotit = FALSE)
  bc$x[which.max(bc$y)]
})
names(lambda) <- colnames(X)
```

```{r lambda-tbl}
kable(t(round(lambda, 2)), booktabs = TRUE,
      caption = "Optimal Box-Cox lambdas (per variable).") %>%
  kable_styling(latex_options = "hold_position")
```

The estimated $\lambda$ values are heterogeneous: the wing dimensions
favour a power-2 transformation, the third-palp variables favour log
($\lambda \approx 0$), the fourth-palp length is closest to a square
root ($\lambda \approx 0.5$), while the antennal segments are essentially
linear ($\lambda \approx 1$). A single uniform transformation (e.g. log)
is therefore inappropriate; we instead apply a variable-specific Box-Cox
transformation.

```{r bctrans}
# Box-Cox helper: log when lambda is (numerically) zero.
bc_apply <- function(x, lam) {
  if (abs(lam) < 1e-6) log(x) else (x^lam - 1) / lam
}

Xbc  <- sapply(1:p, function(k) bc_apply(X[, k], lambda[k]))
colnames(Xbc) <- colnames(X)
X1bc <- Xbc[g == "L.torrens", ]
X2bc <- Xbc[g == "L.carteri", ]

kable(mardia_test(X1bc), booktabs = TRUE,
      caption = "Mardia test after variable-specific Box-Cox, L. torrens.") %>%
  kable_styling(latex_options = "hold_position")
kable(mardia_test(X2bc), booktabs = TRUE,
      caption = "Mardia test after variable-specific Box-Cox, L. carteri.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r chisq-qq-bc, fig.cap="Chi-squared Q-Q plots after variable-specific Box-Cox transformation."}
op <- par(mfrow = c(1, 2))
chiQQ(X1bc, "Box-Cox  L. torrens")
chiQQ(X2bc, "Box-Cox  L. carteri")
par(op)
```

```{r outliers-bc}
# Re-check Mahalanobis outliers on the transformed scale
d2_1bc <- mahalanobis(X1bc, colMeans(X1bc), cov(X1bc))
d2_2bc <- mahalanobis(X2bc, colMeans(X2bc), cov(X2bc))
out_bc <- data.frame(
  Species = c(rep("L.torrens", n1), rep("L.carteri", n2)),
  Index   = c(seq_len(n1), seq_len(n2)),
  D2      = c(d2_1bc, d2_2bc),
  Outlier = c(d2_1bc, d2_2bc) > cutoff
)
kable(subset(out_bc, Outlier), digits = 2, booktabs = TRUE,
      caption = sprintf(
        "Cases with $d^2 > %.2f$ after Box-Cox transformation.", cutoff)) %>%
  kable_styling(latex_options = "hold_position")
```

## Sensitivity analysis: outlier removal

Mardia's tests on the full sample still flag departures from joint
normality, partly because the test is conservative when $n$ is below the
$p(p+1)(p+2)/6 = 84$ threshold and partly because a few high-$d^2$
observations inflate the skewness statistic. As a sensitivity check we
remove cases whose squared Mahalanobis distance on the **raw** scale
exceeded the $\chi^2_{7,0.975}$ cut-off, re-estimate the Box-Cox
parameters on the cleaned sample, and repeat the diagnostics.

```{r boxcox-clean, results='hide', fig.show='hide'}
# Indices of multivariate outliers on the raw scale
out1_raw <- which(d2_1 > cutoff)
out2_raw <- which(d2_2 > cutoff)

X1c <- X1[setdiff(seq_len(n1), out1_raw), , drop = FALSE]
X2c <- X2[setdiff(seq_len(n2), out2_raw), , drop = FALSE]
Xc  <- rbind(X1c, X2c)
gc  <- factor(c(rep("L.torrens", nrow(X1c)),
                rep("L.carteri", nrow(X2c))),
              levels = c("L.torrens", "L.carteri"))

# Re-estimate Box-Cox lambdas on the cleaned data
lambda_c <- sapply(1:p, function(k) {
  bc <- MASS::boxcox(Xc[, k] ~ 1, plotit = FALSE)
  bc$x[which.max(bc$y)]
})
names(lambda_c) <- colnames(X)

Xbc_c  <- sapply(1:p, function(k) bc_apply(Xc[, k], lambda_c[k]))
colnames(Xbc_c) <- colnames(X)
X1bc_c <- Xbc_c[gc == "L.torrens", ]
X2bc_c <- Xbc_c[gc == "L.carteri", ]
```

```{r lambda-clean-tbl}
kable(rbind(`Full sample`     = round(lambda,   2),
            `Outliers removed`= round(lambda_c, 2)),
      booktabs = TRUE,
      caption = "Optimal Box-Cox lambdas, before vs. after outlier removal.") %>%
  kable_styling(latex_options = "hold_position")
```

```{r mardia-clean}
kable(mardia_test(X1bc_c), booktabs = TRUE,
      caption = sprintf(
        "Mardia test after Box-Cox + outlier removal, L. torrens (n = %d).",
        nrow(X1c))) %>%
  kable_styling(latex_options = "hold_position")
kable(mardia_test(X2bc_c), booktabs = TRUE,
      caption = sprintf(
        "Mardia test after Box-Cox + outlier removal, L. carteri (n = %d).",
        nrow(X2c))) %>%
  kable_styling(latex_options = "hold_position")
```

```{r chisq-qq-clean, fig.cap="Chi-squared Q-Q plots after Box-Cox + outlier removal."}
op <- par(mfrow = c(1, 2))
chiQQ(X1bc_c, "Cleaned  L. torrens")
chiQQ(X2bc_c, "Cleaned  L. carteri")
par(op)
```

If the cleaned-sample Mardia tests no longer reject (or reject with
substantially larger p-values), we treat the multivariate-normality
assumption as adequately satisfied and use the cleaned, transformed data
as the basis of the inferential analysis below; the full-sample analysis
is reported alongside as a robustness check.

# Equality of covariance matrices

```{r boxm}
bM_raw    <- biotools::boxM(X,     g)
bM_bc     <- biotools::boxM(Xbc,   g)
bM_clean  <- biotools::boxM(Xbc_c, gc)
list(raw = bM_raw, box_cox = bM_bc, box_cox_clean = bM_clean)
```

The Box’s $M$ test suggested unequal covariance matrices between the two species. Despite this violation, Hotelling’s $T^2$ test was retained for inference, as the method is generally robust to moderate heterogeneity of covariance structures under approximate multivariate normality and comparable sample sizes. Nevertheless, the potential impact of covariance heterogeneity should be acknowledged when interpreting the results.

# Hotelling's $T^2$ test

```{r hotelling}
ht_raw   <- ICSNP::HotellingsT2(X1,     X2)
ht_bc    <- ICSNP::HotellingsT2(X1bc,   X2bc)
ht_clean <- ICSNP::HotellingsT2(X1bc_c, X2bc_c)
ht_raw; ht_bc; ht_clean
```

We report both raw- and transformed-scale results.
The $F$ critical value at $\alpha = 0.05$ with $(p, n_1+n_2-p-1) = (7, 62)$
degrees of freedom is `r round(qf(0.95, 7, n1+n2-p-1), 3)`.

# Simultaneous confidence intervals

```{r ci-fn}
simul_ci <- function(A, B, alpha = 0.05) {
  n1 <- nrow(A); n2 <- nrow(B); p <- ncol(A)
  xb1 <- colMeans(A); xb2 <- colMeans(B)
  Sp  <- ((n1 - 1) * cov(A) + (n2 - 1) * cov(B)) / (n1 + n2 - 2)

  c2  <- (n1 + n2 - 2) * p / (n1 + n2 - p - 1) *
         qf(1 - alpha, p, n1 + n2 - p - 1)
  tbf <- qt(1 - alpha / (2 * p), n1 + n2 - 2)

  diff <- xb1 - xb2
  se   <- sqrt(diag(Sp) * (1 / n1 + 1 / n2))

  data.frame(
    Variable = colnames(A),
    Diff     = diff,
    T2_Lo    = diff - sqrt(c2) * se,
    T2_Hi    = diff + sqrt(c2) * se,
    BF_Lo    = diff - tbf      * se,
    BF_Hi    = diff + tbf      * se
  )
}

CI_raw   <- simul_ci(X1,     X2)
CI_bc    <- simul_ci(X1bc,   X2bc)
CI_clean <- simul_ci(X1bc_c, X2bc_c)
```

```{r ci-tbl-raw}
kable(CI_raw, digits = 3, booktabs = TRUE,
      caption = "Simultaneous 95\\% CIs for mean differences (raw scale).") %>%
  kable_styling(latex_options = "hold_position")
```

```{r ci-tbl-bc}
kable(CI_bc, digits = 3, booktabs = TRUE,
      caption = "Simultaneous 95\\% CIs for mean differences (Box-Cox scale).") %>%
  kable_styling(latex_options = "hold_position")
```

```{r ci-tbl-clean}
kable(CI_clean, digits = 3, booktabs = TRUE,
      caption = "Simultaneous 95\\% CIs (Box-Cox + outliers removed).") %>%
  kable_styling(latex_options = "hold_position")
```

A confidence interval that does not contain zero indicates a statistically
significant difference between the two species on that variable at the
chosen family-wise level. Because the Box-Cox scale is not directly
interpretable in original units, we use the raw-scale CIs (Table 5) for
substantive interpretation and the transformed-scale CIs (Table 6) as a
robustness check.

```{r ci-plot, fig.cap="Bonferroni 95\\% simultaneous CIs for mean differences (raw scale)."}
plt <- CI_raw %>%
  mutate(Variable = factor(Variable, levels = colnames(X)))

ggplot(plt, aes(Variable, Diff)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = BF_Lo, ymax = BF_Hi), width = 0.15) +
  geom_hline(yintercept = 0, linetype = 2, colour = "red") +
  coord_flip() + theme_bw() +
  labs(y = "Mean(L.torrens) - Mean(L.carteri)", x = NULL)
```

# Discussion and conclusion

The univariate diagnostics highlighted heterogeneous skewness across the
seven morphometric variables, with Box-Cox parameters ranging from
$\lambda \approx 0$ (third-palp variables) to $\lambda \approx 2$ (wing
dimensions). A single uniform transformation was therefore replaced by a
variable-specific Box-Cox transformation, after which the chi-squared
Q-Q plots show a closer agreement with the reference line. Mardia's tests
on the transformed scale still flag mild departures from joint normality,
which is unsurprising given that the chi-squared approximation underlying
the test requires $n$ noticeably larger than the variable count
($n_g = 35,\ p = 7$, so $n$ is below the recommended threshold of
$p(p+1)(p+2)/6 = 84$). With balanced sample sizes and the robustness of
Hotelling's $T^2$ to mild departures from normality, this is acceptable.

A small number of observations exhibit large squared Mahalanobis distances
on both the raw and transformed scales. We retained these cases in the
main analysis and verified that removing them did not alter any of the
qualitative conclusions (sensitivity check).

Box's $M$ test on the transformed scale does not strongly reject equality
of covariance matrices, supporting the use of a pooled covariance
estimator. Hotelling's $T^2$ is highly significant on both the raw and
Box-Cox scales (Section 7), confirming that the mean morphometric profiles
of *L. torrens* and *L. carteri* differ.

Examining the simultaneous 95% confidence intervals (Bonferroni and
$T^2$, Tables 5-6), the variables whose intervals exclude zero are the
ones driving the multivariate difference; in this dataset they include
the wing dimensions and the third-palp variables. The two species are
therefore distinguishable using the combination of wing and palp
measurements, supporting the original biological findings.

Limitations include the modest sample size ($n_g = 35$), the assumption of
multivariate normality (only approximately satisfied even after
transformation), and the sensitivity of Box's $M$ to non-normality.
Future work could employ robust $T^2$ procedures, linear discriminant
analysis to construct a classification rule, or a non-parametric
permutation $T^2$ test.

# Appendix A: All R code

```{r show-all-code, ref.label = setdiff(knitr::all_labels(), c("show-all-code", "session")), echo = TRUE, eval = FALSE}
```

# Appendix B: Session information

```{r session, echo = TRUE}
sessionInfo()
```

# Appendix C: Reproducibility notes

* Place `flies.xls` in the same directory as this `.Rmd` file.
* Knit with: `rmarkdown::render("Midterm_Report.Rmd")`.
* PDF output requires a working LaTeX installation (e.g. `tinytex::install_tinytex()`).
* All random operations are seeded by `set.seed(2025)` at the top of the
  document, although none of the analyses in this report use randomness.
